---
title: "Assignment 1"
author: "Ivan Fernandez"
date: "October 13, 2015"
output: html_document
---

# **Reproducible Research**


##Set directory and link raw data

```{r}
  setwd("C:/Users/Ivan/Desktop/datasciencecoursera/Reporesearch")
  data <- read.csv("activity.csv",na.strings="NA")
```


## *What is mean total number of steps taken per day?*
```{r}
  tidydata<-na.omit(data)
         summary(data$steps)

```


```{r}
#Group by day
AggregateDays<-aggregate(steps ~ date, tidydata, sum)

#Produce Hist
hist(AggregateDays$steps, main="Total number of steps per day", xlab="Steps per Day")
```


```{r}
#Calculate mean and median
mean<-mean(AggregateDays$steps)
    median<-median(AggregateDays$steps)

mean

median
```



##*What is the average daily activity pattern?*

```{r}
     #Calculate the average steps with interval.
Interval<-unique(tidydata$interval)
   AverageSteps<-tapply(tidydata$steps, tidydata$interval, mean)

#Make plot of type=1.
plot(Interval,AverageSteps,type="l",ylab="Average steps", main="Average steps per interval")
```


###*Which 5-minute interval, on average across all the days in the dataset contains the maximum number of steps?*

```{r}
MaximumSteps<-names(which.max(AverageSteps))

MaximumSteps

```



#Imputing missing values

##*Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)*
```{r}

dif<-nrow(data)-nrow(tidydata)

dif

```


```{r}
###Create new set of data including na values. Create second table from original data table and plug in average value where steps is na
  nadata <- is.na(data)
    nonadata <- data
      nonadata$steps [which(nadata)] <- AverageSteps[match (data$interval [which (nadata)],data$interval)]

```


```{r}
###re-create first question with new data
   nonaAggregateDays <- aggregate(steps ~ date, nonadata, sum)
      hist(nonaAggregateDays$steps, main="Total Daily Steps", xlab="Daily Steps")
   
```


```{r}
###Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
      
      mean<-mean(nonaAggregateDays$steps)
median<-median(nonaAggregateDays$steps)

###No na
mean
```


```{r}
###No na
median
```
##*Are there differences in activity patterns between weekdays and weekends?*

```{r}
###Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

nonadata$day[weekdays(as.Date(nonadata$date)) %in% c("Saturday", "Sunday")] <- "WEEKEND"
    nonadata$day[!weekdays(as.Date(nonadata$date)) %in% c("Saturday", "Sunday")] <- "WEEKDAY"
        nonadata[, 4] <- as.factor(nonadata[, 4])

```


```{r}
###load packages
library(plyr)
library(ggplot2)
library(lattice)

finalavg <- ddply(nonadata, .(interval, day), summarize, steps = mean(steps))
    xyplot(steps ~ interval | day, data = finalavg, type = "l", layout = c(1, 2), 
        xlab = "5-Min Interval", ylab = "# of steps", main = "Avg Daily Steps")
```